### 2021-05-10 content/80.formal-definition.tex
More LaTeX IsN'tiT

# Appendix

## Formalizing RO-Crate in First Order Logic

@@

    𝕃ro-crate = { Property(p), Class(c), Literal(x), ℝ, 𝕊 }
    𝔻 =  𝕀𝕣𝕚
    𝕀𝕣𝕚 ≡  { IRIs as defined in RFC3987 }
    ℝ ≡  { real or integer numbers }
    𝕊 ≡  { literal strings }

The domain of discourse is the set of `𝕀𝕣𝕚` identifiers [@doi:10.17487/rfc3987] (notation `<http://example.com/>`), with additional descriptions using numbers ℝ (notation `13.37`) and literal strings `𝕊` (notation `“Hello”`).

From this formalized language `𝕃ro-crate` an RO-Crate can be interpreted in any representation that can gather these descriptions, their properties, classes, and literal attributes.

### Minimal RO-Crate

Below is using `𝕃ro-crate` to define a minimal RO-Crate:

    RO-Crate(R) ⊨  Root(R) ∧ Mentions(R, R) ∧
        hasPart(R, d) ∧ Mentions(R, d) ∧ DataEntity(d) ∧
        Mentions(R, c) ∧ ContextualEntity(c)
    ∀r Root(r) →  Dataset(r) ∧ name(r, n)∧ description(r, d)
                            ∧ published(r, date) ∧ license(e, l)
    ∀e∀n name(e, n) →  Literal(n)
    ∀e∀d description(e, d) →  Literal(d)
    ∀e∀date datePublished(e, date) →  Literal(date)
    ∀e∀l license(e, l) →  ContextualEntity(l)
    DataEntity(e) ≡  File(e) ⊕ Dataset(e)
    Entity(e) ≡  DataEntity(e) ∨ ContextualEntity(e)
    ∀e Entity(e) → Class(e)
    Mentions(R, s) ⊨  Relation(s, p, e)  ⊕  Attribute(s, p, l)
    Relation(s, p, o) ⊨ Entity(s) ∧ Property(p) ∧ Entity(o)
    Attribute(s, p, x) ⊨ Entity(s)  ∧ Property(p) ∧ Literal(x)
    Literal(x) ≡  x ∈ ℝ  ⊕   x ∈ 𝕊

An `RO-Crate(R)` is defined as a self-described _Root Data Entity_, which describes and contains parts (_data entities_), which are further described in _contextual entities_.  These terms align with their use in the [RO-Crate 1.1 terminology](https://www.researchobject.org/ro-crate/1.1/terminology).

@@

For simplicity in this formalization (and to assist production rules below) `R` is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. `R` is used by `Mentions(R, e)` to indicate that `e` is an Entity described by the RO-Crate and therefore its metadata (a set of Relation and Attribute predicates) form part of the RO-Crate serialization. `Relation(s, p, o)` and `Attribute(s, p, x)` are defined as a _subject-predicate-object_ triple pattern from an `Entity(s)` using a `Property(p)` to either another `Entity(o)` or a `Literal(x)` value.

### Example of formalized RO-Crate

The below is an example RO-Crate represented using the above formalization, assuming a base URI of `http://example.com/ro/123/`:

@@

In reality many additional attributes from schema.org types like <http://schema.org/Dataset> and <http://schema.org/CreativeWork> would be used to further describe the RO-Crate and its entities, but as these are optional they do not form part of this formalization.

@@

    Mentions(R, s) ∧ Relation(s, p, o) →  Mentions(R, o)
    i ∈ 𝕀𝕣𝕚 → "i"
    r ∈ ℝ →  r
    s ∈ 𝕊 → "s"
    ∀s∀p∀o Relation(s,p,o) →  { "@id": s,
                                p: { "@id": o }
                              }     
    ∀s∀p∀v Attribute(s,p,v) →  { "@id": s,
                                p: v 
                               }
    ∀r∀c  RO-Crate(r) →  { "@graph": [ Mentions(r, c)* ] }
    R ⊨  <./>
    MetadataFileDescriptor(<ro-crate-metadata.json>)

>>>>

%% TODO: Convert to LaTex math mode. See GitHub rendering for original Unicode symbols
% https://github.com/stain/ro-crate-paper/blob/master/content/80.formal-definition.md

% Text remains for now as Markdown

\begin{markdown}
## Formalizing RO-Crate in First Order Logic

@@

\end{markdown}

    $\mathbb{L}$ro-crate = { Property(p), Class(c), Literal(x), ℝ, $\mathbb{S}$ }
    $\mathbb{D}$ =  $\mathbb{Iri}$
    $\mathbb{Iri}$ ≡  { IRIs as defined in RFC3987 }
    ℝ ≡  { real or integer numbers }
    $\mathbb{S}$ ≡  { literal strings }

\begin{markdown}
The domain of discourse is the set of $\mathbb{Iri}$ identifiers [@doi:10.17487/rfc3987] (notation `<http://example.com/>`), with additional descriptions using numbers ℝ (notation `13.37`) and literal strings $\mathbb{S}$ (notation `“Hello”`).

From this formalized language `$\mathbb{L}$ro-crate` an RO-Crate can be interpreted in any representation that can gather these descriptions, their properties, classes, and literal attributes.  
\end{markdown}

\begin{markdown}
### Minimal RO-Crate

Below is using `$\mathbb{L}$ro-crate` to define a minimal RO-Crate:
\end{markdown}

    RO-Crate(R) ⊨  Root(R) ∧ Mentions(R, R) ∧
        hasPart(R, d) ∧ Mentions(R, d) ∧ DataEntity(d) ∧
        Mentions(R, c) ∧ ContextualEntity(c)
    ∀r Root(r) →  Dataset(r) ∧ name(r, n)∧ description(r, d)
                            ∧ published(r, date) ∧ license(e, l)
    ∀e∀n name(e, n) →  Literal(n)
    ∀e∀d description(e, d) →  Literal(d)
    ∀e∀date datePublished(e, date) →  Literal(date)
    ∀e∀l license(e, l) →  ContextualEntity(l)
    DataEntity(e) ≡  File(e) ⊕ Dataset(e)
    Entity(e) ≡  DataEntity(e) ∨ ContextualEntity(e)
    ∀e Entity(e) → Class(e)
    Mentions(R, s) ⊨  Relation(s, p, e)  ⊕  Attribute(s, p, l)
    Relation(s, p, o) ⊨ Entity(s) ∧ Property(p) ∧ Entity(o)
    Attribute(s, p, x) ⊨ Entity(s)  ∧ Property(p) ∧ Literal(x)
    Literal(x) ≡  x ∈ ℝ  ⊕   x ∈  $\mathbb{S}$

\begin{markdown}
An `RO-Crate(R)` is defined as a self-described _Root Data Entity_, which describes and contains parts (_data entities_), which are further described in _contextual entities_.  These terms align with their use in the [RO-Crate 1.1 terminology](https://www.researchobject.org/ro-crate/1.1/terminology).

@@

For simplicity in this formalization (and to assist production rules below) `R` is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. `R` is used by `Mentions(R, e)` to indicate that `e` is an Entity described by the RO-Crate and therefore its metadata (a set of Relation and Attribute predicates) form part of the RO-Crate serialization. `Relation(s, p, o)` and `Attribute(s, p, x)` are defined as a _subject-predicate-object_ triple pattern from an `Entity(s)` using a `Property(p)` to either another `Entity(o)` or a `Literal(x)` value.
\end{markdown}

\begin{markdown}
### Example of formalized RO-Crate

The below is an example RO-Crate represented using the above formalization, assuming a base URI of `http://example.com/ro/123/`:
\end{markdown}

@@

\begin{markdown}
In reality many additional attributes from schema.org types like <http://schema.org/Dataset> and <http://schema.org/CreativeWork> would be used to further describe the RO-Crate and its entities, but as these are optional they do not form part of this formalization.
\end{markdown}

@@

    Mentions(R, s) ∧ Relation(s, p, o) →  Mentions(R, o)
    i ∈ $\mathbb{Iri}$ → "i"
    r ∈ ℝ →  r
    s ∈ $\mathbb{S}$ → "s"
    ∀s∀p∀o Relation(s,p,o) →  { "@id": s,
                                p: { "@id": o }
                              }     
    ∀s∀p∀v Attribute(s,p,v) →  { "@id": s,
                                p: v 
                               }
    ∀r∀c  RO-Crate(r) →  { "@graph": [ Mentions(r, c)* ] }
    R ⊨  <./>
    MetadataFileDescriptor(<ro-crate-metadata.json>)
